---
title: "Class 10: Halloween Mini-Project"
author: "A16125573"
format: pdf
---

# 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
dim(candy)
```

There are 85 rows so 85 candy types

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

38 are fruity candy types

# 2. What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it's winpercent value?

```{r}
rownames(candy)
candy["M&M's", ]$winpercent
```

My favorite candy is M&Ms and their winpercent is 66%

> Q4. What is the winpercent value for "Kit Kat"?

```{r}
candy["Kit Kat", ]$winpercent
```

76 percent

> Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

49 percent

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The variables sugarpercent, pricepercent, and winpercent have different scales as the majority of other columns are categorical (i.e. 0s and 1s)

> Q7. What do you think a zero and one represent for the candy\$chocolate column?

A zero would mean the candy types does not contain chocolate and a one would mean it does

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

> Q9. Is the distribution of winpercent values symmetrical?

It is slightly skewed to the right

> Q10. Is the center of the distribution above or below 50%?

The center is below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choco <- candy$winpercent[as.logical(candy$chocolate)]
fruity <- candy$winpercent[as.logical(candy$fruity)]

mean(choco) > mean(fruity)
```

Chocolate candies are ranked higher

> Q12. Is this difference statistically significant?

```{r}
t.test(choco, fruity)
```

This difference is statistically significant as the p-value is lower than 0.05

# 3. Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
```

```{r}
head(candy[order(candy$winpercent),], n=5)

candy %>% arrange(winpercent) %>% head(5)

```

I prefer the dplyr as it is easier to understand

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(winpercent) %>% tail(5)
```

Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter cup

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

# 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Tootsie Roll Midgies

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The top 5 are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. Nik L Nip is the least popular

# 5. Exploring the correlation structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot, what two variables are anti-correlated (i.e. have minus values)?

fruity and chocolate

> Q23. Similarly, what two variables are most positively correlated?

winpercent and chocolate, as well as chocolate and bar

# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```
```{r}
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard, and pluribus. This does make sense as fruity candies usually do not contain chocolate, caramel etc, most other candies in the dataset are not hard and come in individual packages. 




