---
title: "Class 9 continued"
author: "A16125573"
format: html
---

# 4. Comparative structure analysis of Adenylate Kinase


```{r}
# Install packages in the R console NOT your Rmd/Quarto file

#install.packages("bio3d")
#install.packages("devtools")
#install.packages("BiocManager")

#BiocManager::install("msa")
#devtools::install_bitbucket("Grantlab/bio3d-view")
```
> Q10. Which of the packages above is found only on BioConductor and not CRAN? 

msa

> Q11. Which of the above packages is not found on BioConductor or CRAN?: 

bio3d-view

> Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?

True. 


We will then use this to find all ADK strucutres in the PDB. 

```{r}
library(bio3d)
aa <- get.seq("1ake_A") #1
aa
```

> Q13. How many amino acids are in this sequence, i.e. how long is this sequence?

214 amino acids


We can now run BLAST with this sequence 
```{r}
# Blast or hmmer search 
#b <- blast.pdb(aa) #2

```


```{r}
# Plot a summary of search results
#hits <- plot(b) #3
```

The black dots are the ones we want to analyze.

Let's see what is in the `hits` object. 

```{r}
# List out some 'top hits'
#hits$pdb.id

hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```


```{r}
# Download related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
#split: split it by chain
```

Now I'm going to align and superpose these structures 

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```


```{r}
#pdbs
```

## Principal component analysis

Now we have our aligned and superposed structures, we can perform analysis on them. Let's do PCA

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```



We can cluster the structures by RMSD (root mean squared distance), which gives the sum of all distances 

```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3) #they can be dissected into 3 main groups 

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

# 5. Optional further visualization

Make a movie / trajectory of the major differences of ADK

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```







